Nix for Polyglot, Reproducible Data Science Workflows

### 2025-10-27 paper.qmd
I -> We

---
title: "Nix for Polyglot, Reproducible Data Science Workflows"
format:
    jss-pdf:
        keep-tex: true
    jss-html:
        embed-resources: true
        toc: true
author:
  - name: Bruno Rodrigues
    affiliations:
      - name: Ministry of Research and Higher education, Luxembourg
        department: Department of Statistics
        address: 18, Montée de la Pétrusse
        city: Luxembourg
        country: Luxembourg
        postal-code: 2327
      - Journal of Statistical Software
    orcid: 0000-0002-3211-3689
    email: bruno@brodrigues.co
    url: https://www.brodrigues.co
abstract: |
  Reproducible analysis requires more than clean, well-documented code; it
  demands the precise management of software dependencies, computational
  environments, and workflow execution. While researchers can combine tools like
  [Docker]{.pkg}, [renv]{.pkg} for [R]{.proglang} packages (or [uv]{.pkg} for
  [Python]{.proglang} packages) and [Make]{.pkg} to address these needs, the
  resulting toolchain is often complex and fragile, requiring the coordination
  of multiple disparate systems. This paper introduces a unified framework built
  on the [Nix]{.pkg} package manager, made accessible through two [R]{.proglang}
  packages. The first, [rix]{.pkg}, generates declarative [Nix]{.pkg}
  expressions to define version-pinned, reproducible environments that encompass
  [R]{.proglang}, [Python]{.proglang}, [Julia]{.proglang}, and system-level
  dependencies. The second, [rixpress]{.pkg} (and its [Python]{.proglang} port,
  [ryxpress]{.pkg}), leverages these environments to orchestrate polyglot
  pipelines where each computational step runs in a hermetically sealed,
  language-specific environment with automatic caching. By treating the entire
  computational environment as code, this integrated approach ensures a high
  degree of reproducibility, supports collaboration across heterogeneous
  systems, and helps guarantee that analyses remain executable long into the
  future.

@@

To make [Nix]{.pkg} accessible to researchers, I developed two [R]{.proglang}
packages:

@@

To make these capabilities more approachable for [R]{.proglang} users seeking
reproducible, project-specific environments without learning the full
[Nix]{.proglang} language, I created the [rix]{.pkg} and [rixpress]{.pkg}
packages.

@@

This expression can be saved in a file called `default.nix`. The environment can
then be built on a system with [Nix]{.pkg} installed using the `nix-build`
command.^[For installing [Nix]{.pkg}, I recommend the Determinate Systems
installer: \url{https://determinate.systems/posts/determinate-nix-installer}].
Once the build completes, the user can enter the interactive shell with
`nix-shell`, which executes following steps:

@@

I strive to provide an available date per week: each Monday, a GitHub Action
tests popular packages on Linux and macOS, and only if all tests succeed is the
date added to the list of available dates in [rix]{.pkg}. This ensures users can
reliably install packages, and allows me to backport fixes if needed. For
example, when RStudio was temporarily broken due to a dependency issue
(`boost`), a pull request was submitted to the official `nixpkgs` repository. I
backported the fix to the `rstats-on-nix` fork, making RStudio available to
users of [rix]{.pkg} earlier than upstream, as merging PRs in the official
repository can take some time.

I have backported fixes to the `rstats-on-nix` `nixpkgs` fork as far back as
March 2019. The process involves checking out a `nixpkgs` commit on the selected
date, updating the [R]{.proglang} package set using Posit CRAN and Bioconductor
snapshots, backporting fixes, and ensuring popular packages work on both
x86-linux (including WSL2) and aarch64-darwin (Apple Silicon). These changes are
committed to a dated branch in `rstats-on-nix/nixpkgs`. Users can see all
available dates with `rix::available_dates()`.

A drawback of forking `nixpkgs` is that backported packages are not included
upstream and thus are not prebuilt by Nix’s CI platform, Hydra. Users may need
to build many packages from source, which can be time-consuming. To mitigate
this, I provide a binary cache sponsored by [Cachix](https://www.cachix.org/),
complementing the public Nix cache. Instructions for using Cachix are in
[rix]{.pkg}’s documentation. Using the cache significantly speeds up
installations, as prebuilt packages are downloaded rather than compiled.

@@

While the reliance on the `rstats-on-nix` fork might initially raise
sustainability concerns, this risk should be contextualized. The fork serves a
focused purpose: it updates CRAN and [Julia]{.proglang} package registries daily
and backports fixes when builds break. Importantly, users are never locked into
this fork—they retain the ability to point directly to any commit in the
upstream `nixpkgs` repository, maintaining full access to the official
[Nix]{.pkg} ecosystem. The fork introduces no fundamental architectural changes
or proprietary extensions; it functions as a convenience layer providing tested
package sets and dated snapshots. Should maintenance cease (which is unlikely,
as I directly use it myself), users could seamlessly transition to upstream
`nixpkgs` commits, though with trade-offs: less frequent [R]{.proglang} and
[Julia]{.proglang} package updates. The fork's value lies not in creating
dependency, but in reducing friction through testing infrastructure and curated
"known-good" dates. Thus, the fork represents an enhancement to user experience
rather than a critical single point of failure—the underlying [Nix]{.pkg}
infrastructure and official `nixpkgs` repository remain the foundation, ensuring
long-term viability of environments created with [rix]{.pkg}.

@@

To illustrate these capabilities with a realistic research task, I present a
polyglot pipeline that simulates a Real Business Cycle (RBC) model in
[Julia]{.proglang}, uses the resulting data to train an
[XGBoost]{.pkg} forecasting model in [Python]{.proglang}, visualises the results
in [R]{.proglang} with [ggplot2]{.pkg}, and finally compiles a [Quarto]{.pkg}
report. The code for this example can be found in the [following
repository](https://github.com/b-rodrigues/rixpress_demos/tree/master/rbc)^[https://github.com/b-rodrigues/rixpress_demos/tree/master/rbc],
with additional details provided in the Appendix of this manuscript.

@@

However, solving such a complex problem makes [Nix]{.pkg} a complex tool. With
[rix]{.pkg}, I aim to make [Nix]{.pkg} more accessible to [R]{.proglang} users
by providing a familiar interface and workflow. By building reproducible
development shells with [Nix]{.pkg}, researchers can accommodate a wide range of
use cases: running scripts and pipelines, developing interactive [shiny]{.pkg}
applications, or serving [plumber]{.pkg} APIs.

@@

I thank the rOpenSci reviewers and contributors who provided valuable feedback
on the development of [rix]{.pkg} and [rixpress]{.pkg}. In particular, I am
grateful to David Watkins and Jacob Wujiciak-Jens for their reviews of
[rix]{.pkg}, and to William Landau and Anthony Martinez for their reviews of
[rixpress]{.pkg}. I also acknowledge the contributions of Philipp Baumann,
[rix]{.pkg}’s co-author, Richard J. Acton, Jordi Rosell, Elio Campitelli, László
Kupcsik, and Michael Heming for [rix]{.pkg}. Their expertise and feedback
greatly improved the quality and usability of these packages. I would also like
to thank Pol Dellaiera and Edvin Syk for providing feedback on the manuscript.

>>>>

---
title: "Nix for Polyglot, Reproducible Data Science Workflows"
format:
    jss-pdf:
        keep-tex: true
    jss-html:
        embed-resources: true
        toc: true
author:
  - name: Bruno Rodrigues
    affiliations:
      - name: Ministry of Research and Higher education, Luxembourg
        department: Department of Statistics
        address: 18, Montée de la Pétrusse
        city: Luxembourg
        country: Luxembourg
        postal-code: 2327
      - Journal of Statistical Software
    orcid: 0000-0002-3211-3689
    email: bruno@brodrigues.co
    url: https://www.brodrigues.co
  - name: Philipp Baumann
    affiliations:
      - name: Alliance SwissPass, Switzerland
        department: Data and Analytics
        address: 7, Länggassstrasse
        city: Bern
        country: Switzerland
        postal-code: 3012
    orcid: 0000-0002-3194-8975
    email: baumann-philipp@protonmail.com
abstract: |
  Reproducible analysis requires more than clean, well-documented code; it
  demands the precise management of software dependencies, computational
  environments, and workflow execution. While researchers can combine tools like
  [Docker]{.pkg}, [renv]{.pkg} for [R]{.proglang} packages (or [uv]{.pkg} for
  [Python]{.proglang} packages) and [Make]{.pkg} to address these needs, the
  resulting toolchain is often complex and fragile, requiring the coordination
  of multiple disparate systems. This paper introduces a unified framework built
  on the [Nix]{.pkg} package manager, made accessible through two [R]{.proglang}
  packages. The first, [rix]{.pkg}, generates declarative [Nix]{.pkg}
  expressions to define version-pinned, reproducible environments that encompass
  [R]{.proglang}, [Python]{.proglang}, [Julia]{.proglang}, and system-level
  dependencies. The second, [rixpress]{.pkg} (and its [Python]{.proglang} port,
  [ryxpress]{.pkg}), leverages these environments to orchestrate polyglot
  pipelines where each computational step runs in a hermetically sealed,
  language-specific environment with automatic caching. By treating the entire
  computational environment as code, this integrated approach ensures a high
  degree of reproducibility, supports collaboration across heterogeneous
  systems, and helps guarantee that analyses remain executable long into the
  future.

@@

To make [Nix]{.pkg} accessible to researchers, we developed two [R]{.proglang}
packages:

@@

To make these capabilities more approachable for [R]{.proglang} users seeking
reproducible, project-specific environments without learning the full
[Nix]{.proglang} language, we created the [rix]{.pkg} and [rixpress]{.pkg}
packages.

@@

This expression can be saved in a file called `default.nix`. The environment can
then be built on a system with [Nix]{.pkg} installed using the `nix-build`
command.^[For installing [Nix]{.pkg}, we recommend the Determinate Systems
installer: \url{https://determinate.systems/posts/determinate-nix-installer}].
Once the build completes, the user can enter the interactive shell with
`nix-shell`, which executes following steps:

@@

We strive to provide an available date per week: each Monday, a GitHub Action
tests popular packages on Linux and macOS, and only if all tests succeed is the
date added to the list of available dates in [rix]{.pkg}. This ensures users can
reliably install packages, and allows me to backport fixes if needed. For
example, when RStudio was temporarily broken due to a dependency issue
(`boost`), a pull request was submitted to the official `nixpkgs` repository. I
backported the fix to the `rstats-on-nix` fork, making RStudio available to
users of [rix]{.pkg} earlier than upstream, as merging PRs in the official
repository can take some time.

We have backported fixes to the `rstats-on-nix` `nixpkgs` fork as far back as
March 2019. The process involves checking out a `nixpkgs` commit on the selected
date, updating the [R]{.proglang} package set using Posit CRAN and Bioconductor
snapshots, backporting fixes, and ensuring popular packages work on both
x86-linux (including WSL2) and aarch64-darwin (Apple Silicon). These changes are
committed to a dated branch in `rstats-on-nix/nixpkgs`. Users can see all
available dates with `rix::available_dates()`.

A drawback of forking `nixpkgs` is that backported packages are not included
upstream and thus are not prebuilt by Nix’s CI platform, Hydra. Users may need
to build many packages from source, which can be time-consuming. To mitigate
this, we provide a binary cache sponsored by [Cachix](https://www.cachix.org/),
complementing the public Nix cache. Instructions for using Cachix are in
[rix]{.pkg}’s documentation. Using the cache significantly speeds up
installations, as prebuilt packages are downloaded rather than compiled.

@@

While the reliance on the `rstats-on-nix` fork might initially raise
sustainability concerns, this risk should be contextualized. The fork serves a
focused purpose: it updates CRAN and [Julia]{.proglang} package registries daily
and backports fixes when builds break. Importantly, users are never locked into
this fork—they retain the ability to point directly to any commit in the
upstream `nixpkgs` repository, maintaining full access to the official
[Nix]{.pkg} ecosystem. The fork introduces no fundamental architectural changes
or proprietary extensions; it functions as a convenience layer providing tested
package sets and dated snapshots. Should maintenance cease, users could
seamlessly transition to upstream `nixpkgs` commits, though with trade-offs:
less frequent [R]{.proglang} and [Julia]{.proglang} package updates. The fork's
value lies not in creating dependency, but in reducing friction through testing
infrastructure and curated "known-good" dates. Thus, the fork represents an
enhancement to user experience rather than a critical single point of
failure—the underlying [Nix]{.pkg} infrastructure and official `nixpkgs`
repository remain the foundation, ensuring long-term viability of environments
created with [rix]{.pkg}.

@@

To illustrate these capabilities with a realistic research task, we present a
polyglot pipeline that simulates a Real Business Cycle (RBC) model in
[Julia]{.proglang}, uses the resulting data to train an
[XGBoost]{.pkg} forecasting model in [Python]{.proglang}, visualises the results
in [R]{.proglang} with [ggplot2]{.pkg}, and finally compiles a [Quarto]{.pkg}
report. The code for this example can be found in the [following
repository](https://github.com/b-rodrigues/rixpress_demos/tree/master/rbc)^[https://github.com/b-rodrigues/rixpress_demos/tree/master/rbc],
with additional details provided in the Appendix of this manuscript.

@@

However, solving such a complex problem makes [Nix]{.pkg} a complex tool. With
[rix]{.pkg}, we aim to make [Nix]{.pkg} more accessible to [R]{.proglang} users
by providing a familiar interface and workflow. By building reproducible
development shells with [Nix]{.pkg}, researchers can accommodate a wide range of
use cases: running scripts and pipelines, developing interactive [shiny]{.pkg}
applications, or serving [plumber]{.pkg} APIs.

@@

We thank the rOpenSci reviewers and contributors who provided valuable feedback
on the development of [rix]{.pkg} and [rixpress]{.pkg}. In particular, we are
grateful to David Watkins and Jacob Wujiciak-Jens for their reviews of
[rix]{.pkg}, and to William Landau and Anthony Martinez for their reviews of
[rixpress]{.pkg}. We also acknowledge the contributions of Richard J. Acton,
Jordi Rosell, Elio Campitelli, László Kupcsik, and Michael Heming for
[rix]{.pkg}. Their expertise and feedback greatly improved the quality and
usability of these packages. We would also like to thank Pol Dellaiera and Edvin
Syk for providing feedback on the manuscript.
